---
title: "Data visualizations based on Tufte's principles"
subtitle: "Ten rules for the figures in a clinical study report"
author: "15 minutes, for clinical statisticians"
format:
  revealjs:
    theme: [default, theme.scss]
    width: 1280
    height: 720
    margin: 0.07
    center: false
    transition: none
    slide-number: c/t
    hash: true
    embed-resources: true
    title-slide-attributes:
      data-background-color: "#faf8f2"
execute:
  echo: false
  warning: false
  message: false
fig-dpi: 192
---

```{r setup}
source("R/theme_tufte.R")
library(survival)
library(dplyr)
adtte  <- read.csv("data/adtte.csv")
tumor  <- read.csv("data/tumor_wk12.csv")
orr    <- read.csv("data/orr.csv")
lab_means <- read.csv("data/lab_means.csv")
egfr   <- read.csv("data/egfr.csv")
prof   <- read.csv("data/dose_profile.csv")
forest <- read.csv("data/forest.csv")
bor    <- read.csv("data/bor.csv")
pk     <- read.csv("data/pk.csv")
alt_subj <- read.csv("data/alt_subj.csv")

# KM pieces used on several slides
sf <- survfit(Surv(aval, 1 - cnsr) ~ trt, data = adtte)
med <- summary(sf)$table[, "median"]
km <- data.frame(
  time  = sf$time, surv = sf$surv, ncens = sf$n.censor,
  trt   = rep(sub("trt=", "", names(sf$strata)), sf$strata)
)
km0 <- rbind(data.frame(time = 0, surv = 1, ncens = 0,
                        trt = unique(km$trt)), km)
atrisk <- adtte |>
  cross_join(data.frame(month = c(0, 6, 12, 18, 24))) |>
  summarize(n_at_risk = sum(aval >= month), .by = c(trt, month))
lab_means$analyte <- factor(lab_means$analyte,
                            levels = c("ALT", "AST", "ALP", "Bilirubin",
                                       "Creatinine", "Hemoglobin"))
prof$dose <- factor(prof$dose, levels = c("placebo", "10 mg", "20 mg"))
knitr::opts_chunk$set(dev = "ragg_png")
```

## Above all else show the data

::: {.eyebrow}
A working deck [MER-201 and MER-102 are invented studies. The rules are not.]{}
:::

::: {.lede}
Edward Tufte's ten rules for data graphics, each demonstrated on a figure a
clinical statistician actually makes, almost always as a pair: the junk
version on the left, the revised version on the right. Then the kill list,
drawn the same way, and the checks to run before a figure goes in the CSR.
:::

```{r cover, fig.width=11.4, fig.height=3.2}
ggplot(km0, aes(time, surv, group = trt)) +
  geom_step(aes(colour = trt), linewidth = 0.7) +
  scale_colour_manual(values = c(pal$ink, pal$g2)) +
  annotate("text", x = 24.3, y = km0$surv[km0$trt == "active treatment" &
             km0$time == max(km0$time[km0$trt == "active treatment"])][1],
           label = "active treatment", hjust = 0, family = "Charter",
           size = 4, colour = pal$ink) +
  annotate("text", x = 24.3, y = km0$surv[km0$trt == "placebo" &
             km0$time == max(km0$time[km0$trt == "placebo"])][1],
           label = "placebo", hjust = 0, family = "Charter",
           size = 4, colour = pal$g3) +
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24), limits = c(0, 32.5)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) +
  labs(x = "months", y = "PFS") +
  theme_tufte(15)
```

::: {.caption}
MER-201 progression-free survival, N = 240. Two curves, two labels, five ticks.
Everything else on this slide waits for rule 3.
:::

::: notes
0.5 min. Frame the talk: ten rules from three books (VDQI 1983, Envisioning
Information 1990, Visual Explanations 1997). Every figure is from one invented
phase 2 study so the numbers reconcile across slides.
:::

## The reader is a reviewer

::: {.lede}
Your figures face the busiest readers in the industry. A DSMB member with six
studies on the table. A regulator with a 400-page CSR and a PDUFA date.
Every mark that is not data costs them attention they were saving for the data.
:::

:::: {.columns}
::: {.column width="50%"}
|    | Rule | Figure it fixes |
|----|------|-----------------|
| 1  | Show the data | The dynamite plot |
| 2  | Maximize the data-ink ratio | The KM plot, audited |
| 3  | Erase non-data-ink | The default KM plot |
| 4  | Erase redundant data-ink | The ORR bar, said once |
| 5  | Keep graphical integrity | The truncated axis |

: {tbl-colwidths="[8,50,42]"}
:::
::: {.column width="50%"}
|    | Rule | Figure it fixes |
|----|------|-----------------|
| 6  | Small multiples | The six-lab overlay |
| 7  | Layer and separate | The heavy grid |
| 8  | Micro and macro | The spaghetti plot, kept |
| 9  | Smallest effective difference | The three-arm rainbow |
| 10 | Integrate words and data | The forest plot |

: {tbl-colwidths="[8,50,42]"}
:::
::::

::: notes
1 min. Tufte's frame: data-ink over total ink. In pharma the stakes are higher
than aesthetics; a figure that hides patients hides risk.
:::

## Rule 1 · Show the data

::: {.eyebrow}
Rule 01 [VDQI, p. 92 — "Above all else show the data"]{}
:::

:::: {.columns}
::: {.column width="50%"}
[As most tools draw it]{.panel-lab .bad}
```{r dynamite, fig.width=5.2, fig.height=3.9}
arm_means <- tumor |>
  summarize(mean_pchg = mean(pchg),
            se_pchg = sd(pchg) / sqrt(n()), .by = trt)
ggplot(arm_means, aes(trt, mean_pchg, fill = trt)) +
  geom_col(width = 0.55, colour = "grey25") +
  geom_errorbar(aes(ymin = mean_pchg - se_pchg, ymax = mean_pchg + se_pchg),
                width = 0.18, linewidth = 0.7, colour = "grey25") +
  scale_x_discrete(labels = c("active treatment", "placebo")) +
  labs(title = "Mean change in tumor size, week 12",
       x = NULL, y = "% change (mean ± SE)") +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[The same 120 patients]{.panel-lab}
```{r dots, fig.width=5.2, fig.height=3.9}
set.seed(4)
tumor_medians <- summarize(tumor, pchg = median(pchg), .by = trt)
ggplot(tumor, aes(trt, pchg)) +
  geom_hline(yintercept = 0, colour = pal$g2, linewidth = 0.3) +
  geom_jitter(width = 0.16, size = 1.1, alpha = 0.55, colour = pal$ink) +
  geom_errorbar(data = tumor_medians, aes(ymin = pchg, ymax = pchg),
                width = 0.4, linewidth = 0.8, colour = pal$accent) +
  geom_text(data = tumor_medians, aes(label = sprintf("%.0f%%", pchg)),
            hjust = 0, nudge_x = 0.26, family = "Menlo",
            size = 3.4, colour = pal$accent, fontface = "bold") +
  scale_x_discrete(labels = c("active treatment", "placebo")) +
  labs(title = "Change in tumor size, week 12, every patient",
       x = NULL, y = "% change from baseline") +
  theme_tufte(15)
```
:::
::::

::: {.caption}
The bar shows 2 numbers from 120 patients and hides the rest: the spread, the
placebo responders, the active treatment progressors. The dot plot shows all 120
and the medians in red. Same data, opposite amount of information.
:::

::: notes
1.5 min. The dynamite plot is the most common figure in the industry and the
least informative. Point at the placebo patients below -30%: invisible on the
left.
:::

## Rule 2 · Maximize the data-ink ratio

::: {.eyebrow}
Rule 02 [VDQI, p. 96 — data-ink is ink that would change if the data changed]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Non-data-ink, painted red]{.panel-lab .bad}
```{r inkaudit, fig.width=5.2, fig.height=3.9}
ggplot(km0, aes(time, surv, colour = trt)) +
  geom_step(linewidth = 0.7) +
  scale_colour_manual(NULL, values = c(pal$ink, pal$g3),
                      labels = c("active treatment", "placebo")) +
  scale_x_continuous(breaks = 0:12 * 2, minor_breaks = 0:24) +
  scale_y_continuous(breaks = 0:10 / 10) +
  labs(title = "Kaplan-Meier plot of PFS", x = "Time (months)", y = "PFS") +
  theme_minimal(15, base_family = "Charter") +
  theme(
    plot.background  = element_rect(fill = pal$paper, colour = NA),
    panel.grid.major = element_line(colour = pal$accent, linewidth = 0.3),
    panel.grid.minor = element_line(colour = pal$accent, linewidth = 0.15),
    panel.border     = element_rect(colour = pal$accent, fill = NA,
                                    linewidth = 0.8),
    axis.ticks       = element_line(colour = pal$accent, linewidth = 0.5),
    axis.ticks.length = unit(4, "pt"),
    axis.text        = element_text(family = "Menlo", size = 8,
                                    colour = pal$g3),
    axis.title       = element_text(family = "Menlo", size = 9,
                                    colour = pal$muted),
    plot.title       = element_text(family = "Menlo", size = 10,
                                    colour = pal$muted),
    legend.background = element_rect(colour = pal$accent, fill = pal$paper),
    legend.text      = element_text(size = 9),
    legend.position  = c(0.8, 0.85)
  )
```
:::
::: {.column width="50%"}
[Only ink the data can move]{.panel-lab}
```{r inkonly, fig.width=5.2, fig.height=3.9}
curve_ends <- km0 |> slice_tail(n = 1, by = trt)
ggplot(km0, aes(time, surv, group = trt)) +
  geom_step(aes(colour = trt), linewidth = 0.7) +
  scale_colour_manual(values = c(pal$ink, pal$g2)) +
  annotate("text", x = curve_ends$time + 0.5, y = curve_ends$surv,
           label = c("active treatment", "placebo"), hjust = 0,
           family = "Charter", size = 3.6, colour = c(pal$ink, pal$g3)) +
  scale_x_continuous(breaks = c(0, 12, 24), limits = c(0, 33.5)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) +
  labs(x = "months", y = "PFS") +
  theme_tufte(15)
```
:::
::::

::: {.caption}
The audit: change every event time in the dataset and ask what moves. The
curves move. Nothing red does: not the frame, the grid, the minor ticks, the
legend box. The right panel keeps the curves, three ticks per axis, and two
names. The ratio of ink that works to ink that sits went from roughly a third
to nearly all of it.
:::

::: notes
0.5 min. This is the test to run on any figure: would this mark change if the
data changed? Red here is a teaching device; the accent never goes on
furniture in a real figure.
:::

## Rule 3 · Erase non-data-ink

::: {.eyebrow}
Rule 03 [VDQI, p. 96 — border, legend, grid, minor ticks, in that order]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Default output]{.panel-lab .bad}
```{r kmbad, fig.width=5.2, fig.height=3.9}
ggplot(km0, aes(time, surv, colour = trt)) +
  geom_step(linewidth = 1.1) +
  geom_point(data = filter(km0, ncens > 0), shape = 3, size = 1.6) +
  scale_colour_manual(NULL, values = c("#00BFC4", "#F8766D")) +
  scale_x_continuous(breaks = 0:8 * 3) +
  labs(title = "Kaplan-Meier plot of PFS (FAS)",
       x = "Time (months)", y = "Survival probability") +
  theme_junk(15) +
  theme(legend.position = c(0.78, 0.85))
```
:::
::: {.column width="50%"}
[After erasure]{.panel-lab}
```{r kmgood, fig.width=5.2, fig.height=3.9}
ggplot(km0, aes(time, surv, group = trt)) +
  geom_step(aes(colour = trt), linewidth = 0.7) +
  geom_point(data = filter(km0, ncens > 0), aes(colour = trt),
             shape = "|", size = 1.8) +
  scale_colour_manual(values = c(pal$ink, pal$g2)) +
  annotate("segment", x = med[1], xend = med[1], y = 0, yend = 0.5,
           colour = pal$accent, linewidth = 0.35, linetype = "22") +
  annotate("text", x = med[1] + 0.4, y = 0.06, family = "Menlo", size = 3.2,
           colour = pal$accent, hjust = 0,
           label = sprintf("median %.1f mo", med[1])) +
  annotate("text", x = curve_ends$time[1] + 0.5,
           y = curve_ends$surv[1] + 0.03,
           label = "active treatment", hjust = 0, family = "Charter",
           size = 3.6, colour = pal$ink) +
  annotate("text", x = curve_ends$time[2] + 0.5,
           y = curve_ends$surv[2] - 0.01,
           label = sprintf("placebo\nmedian %.1f mo", med[2]), hjust = 0,
           family = "Charter", size = 3.6, colour = pal$g3, lineheight = 0.95) +
  geom_text(data = filter(atrisk, trt == "active treatment"),
            aes(month, -0.24, label = n_at_risk), family = "Menlo",
            size = 2.9, colour = pal$g3) +
  geom_text(data = filter(atrisk, trt == "placebo"),
            aes(month, -0.34, label = n_at_risk), family = "Menlo",
            size = 2.9, colour = pal$g2) +
  annotate("text", x = -1.6, y = c(-0.24, -0.34), hjust = 1, family = "Menlo",
           size = 2.9, colour = c(pal$g3, pal$g2),
           label = c("at risk  act", "plc")) +
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24), limits = c(-3.5, 34.5)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(x = NULL, y = "PFS") +
  theme_tufte(15) +
  theme(plot.margin = margin(8, 8, 44, 8))
```
:::
::::

::: {.caption}
Erasure order: frame box, grid, legend, minor ticks. What replaced them is
data: direct labels, censor marks, medians where they occur, and the at-risk
counts, which are the honest width of the right-hand tail.
:::

::: notes
1.5 min. The legend is the tell: it asks the reader to hold color pairs in
memory. The at-risk table is not decoration; past month 18 both curves stand
on fewer than 35 patients.
:::

## Rule 4 · Erase redundant data-ink

::: {.eyebrow}
Rule 04 [VDQI, p. 100 — one encoding per quantity]{}
:::

:::: {.columns}
::: {.column width="50%"}
[42.3, said five ways]{.panel-lab .bad}
```{r redundant, fig.width=5.2, fig.height=3.9}
ggplot(orr, aes(trt, orr, fill = trt)) +
  geom_col(width = 0.5, colour = "grey25") +
  geom_text(aes(label = sprintf("%.1f%%", orr)), vjust = -0.6, size = 3.6) +
  scale_y_continuous(breaks = 0:9 * 5, limits = c(0, 48)) +
  scale_x_discrete(labels = c("active treatment", "placebo")) +
  labs(title = "Overall response rate", x = NULL, y = "ORR (%)") +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[Position once, decimals printed]{.panel-lab}
```{r once, fig.width=5.2, fig.height=3.9}
ggplot(orr, aes(trt, orr)) +
  geom_point(size = 3, colour = c(pal$accent, pal$ink)) +
  geom_text(aes(label = sprintf("%.1f%%", orr)), vjust = -1.1,
            family = "Menlo", size = 3.6,
            colour = c(pal$accent, pal$ink)) +
  scale_y_continuous(breaks = c(0, 45), limits = c(0, 48)) +
  scale_x_discrete(labels = c("active treatment", "placebo")) +
  labs(title = "Overall response rate (%)", x = NULL, y = NULL) +
  theme_tufte(15)
```
:::
::::

::: {.caption}
The left bar states 42.3 five times: top-edge position, fill height, printed
label, the gridline it touches, the axis tick beside it. Five statements of
one number are not clearer, they are louder. The right panel keeps two: the
dot's position, and the printed value for the reader who needs the decimals.
:::

::: notes
0.5 min. A number plus a sparkline is two encodings doing different work.
A bar plus its own label plus a grid is one encoding shouting.
:::

## Rule 5 · Never let the picture out-run the number

::: {.eyebrow}
Rule 05 [Lie factor = visual effect ÷ data effect. Keep it between 0.95 and 1.05]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Axis starts at 35]{.panel-lab .bad}
```{r orrbad, fig.width=5.2, fig.height=3.6}
ggplot(orr, aes(trt, orr, fill = trt)) +
  geom_col(width = 0.5, colour = "grey25") +
  coord_cartesian(ylim = c(35, 45)) +
  scale_x_discrete(labels = c("active treatment", "placebo")) +
  labs(title = "Overall response rate (%)", x = NULL, y = NULL) +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[Axis starts at zero]{.panel-lab}
```{r orrgood, fig.width=5.2, fig.height=3.6}
ggplot(orr, aes(trt, orr)) +
  geom_col(width = 0.5, fill = c(pal$accent, pal$g2)) +
  scale_x_discrete(labels = c("active treatment", "placebo")) +
  scale_y_continuous(limits = c(0, 48), breaks = c(0, 20, 40)) +
  labs(title = "Overall response rate (%)", x = NULL, y = NULL) +
  theme_tufte(15)
```
:::
::::

::: {.caption}
ORR 42.3% vs 38.1%: an 11% relative difference. The truncated panel draws a
135% visual difference. Lie factor 12. Bars encode length, so their baseline
is zero, and per rule 4 the length is said once: the bars carry the
comparison, this caption carries the decimals. If a difference is only
visible on a truncated axis, that is the finding, and the honest home for it
is the number itself with its CI.
:::

::: notes
1 min. Also on the lie list: response bubbles scaled by radius, subgroup
panels on different scales. The dual-axis overlay gets its own slide in the
kill list.
:::

## Rule 6 · Small multiples

::: {.eyebrow}
Rule 06 [VDQI, p. 67 — "Compared to what?"]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Twelve lines, one panel]{.panel-lab .bad}
```{r labsoverlay, fig.width=5.2, fig.height=3.9}
ggplot(lab_means, aes(week, pchg, colour = analyte, linetype = trt)) +
  geom_line(linewidth = 0.8) +
  scale_linetype_discrete(NULL, labels = c("active treatment", "placebo")) +
  # legend order is not guaranteed across sessions unless pinned
  guides(linetype = guide_legend(order = 1),
         colour = guide_legend(order = 2)) +
  labs(title = "Laboratory parameters over time",
       x = "Week", y = "Mean % change", colour = "Parameter") +
  theme_junk(15) +
  theme(legend.key.size = unit(11, "pt"),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9))
```
:::
::: {.column width="50%"}
[Six panels, one line each]{.panel-lab}
```{r labs, fig.width=5.2, fig.height=3.9}
lab_means <- lab_means |>
  mutate(key = interaction(analyte, trt),
         col = case_when(trt == "placebo" ~ "plc",
                         analyte == "ALT" ~ "focal",
                         .default = "act"))
lab_end <- filter(lab_means, week == 26, trt != "placebo")
arm_labels <- data.frame(
  week = c(11, 15), pchg = c(13, -4.5),
  lab = c("active treatment", "placebo"),
  analyte = factor("ALT", levels = levels(lab_means$analyte)),
  key = NA, col = c("focal", "plc")
)
ggplot(lab_means, aes(week, pchg, group = key, colour = col)) +
  geom_hline(yintercept = 0, colour = pal$g1, linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(data = lab_end, size = 1.3) +
  geom_text(data = arm_labels, aes(label = lab), family = "Charter",
            size = 2.9, hjust = 0) +
  scale_colour_manual(values = c(plc = pal$g2, act = pal$ink,
                                 focal = pal$accent)) +
  facet_wrap(~analyte, nrow = 2) +
  scale_x_continuous(breaks = c(0, 26)) +
  scale_y_continuous(breaks = c(0, 25)) +
  labs(x = "week", y = "mean % change") +
  theme_tufte(15) +
  theme(panel.spacing = unit(10, "pt"))
```
:::
::::

::: {.caption}
Mean % change from baseline, six labs, two arms. On the left the ALT signal is
one strand in a twelve-line knot with a two-part legend. On the right, six
panels on one shared scale, active treatment in ink, placebo in gray; the ALT
climb to +24% is the safety story, so ALT alone carries the accent.
:::

::: notes
1 min. The eye reads parallel panels faster than it untangles overlap. Same
scale in every panel, same order every time the figure repeats in the CSR.
:::

## Rule 7 · Layer and separate

::: {.eyebrow}
Rule 07 [Envisioning Information — data loudest, scaffolding faintest]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Grid at the weight of the data]{.panel-lab .bad}
```{r layerbad, fig.width=5.2, fig.height=3.9}
ggplot(prof, aes(week, pchg, group = dose)) +
  geom_line(linewidth = 1, colour = "grey15") +
  scale_x_continuous(breaks = 0:13 * 2) +
  scale_y_continuous(breaks = -13:1 * 2) +
  labs(title = "Tumor burden by dose (MER-102)",
       x = "Week", y = "Mean % change") +
  theme_grey(15, base_family = "Helvetica") +
  theme(panel.grid.major = element_line(colour = "grey40", linewidth = 0.5),
        panel.grid.minor = element_line(colour = "grey55", linewidth = 0.35),
        panel.border = element_rect(colour = "grey15", fill = NA,
                                    linewidth = 0.9),
        plot.title = element_text(size = 9.5, hjust = 0))
```
:::
::: {.column width="50%"}
[Grid one step back]{.panel-lab}
```{r layergood, fig.width=5.2, fig.height=3.9}
dose_ends <- filter(prof, week == 26)
ggplot(prof, aes(week, pchg, group = dose)) +
  geom_hline(yintercept = c(-20, -10), colour = pal$g1, linewidth = 0.4) +
  geom_line(linewidth = 0.7, colour = pal$ink) +
  geom_text(data = dose_ends, aes(label = dose), hjust = 0, nudge_x = 0.6,
            family = "Charter", size = 3.4, colour = pal$ink) +
  scale_x_continuous(breaks = c(0, 12, 26), limits = c(0, 30.5)) +
  scale_y_continuous(breaks = c(0, -10, -20)) +
  labs(title = "Tumor burden by dose (MER-102)",
       x = "week", y = "mean % change") +
  theme_tufte(15)
```
:::
::::

::: {.caption}
Same three dose groups. On the left the grid is drawn at the weight of the
data and every crossing makes a small visual event Tufte calls 1 + 1 = 3.
On the right the layers are ranked: data in ink, labels next, two faint
gridlines last. Nothing crosses anything at equal weight.
:::

::: notes
0.5 min. The fix for a vibrating chart is almost always to lighten the grid,
not to thicken the data.
:::

## Rule 8 · Micro and macro readings

::: {.eyebrow}
Rule 08 [Envisioning Information — "To clarify, add detail"]{}
:::

:::: {.columns}
::: {.column width="50%"}
[The summary alone]{.panel-lab .bad}
```{r egfrmean, fig.width=5.2, fig.height=3.9}
egfr_means <- egfr |>
  summarize(mean_egfr = mean(egfr),
            se_egfr = sd(egfr) / sqrt(n()), .by = week)
ggplot(egfr_means, aes(week, mean_egfr)) +
  geom_errorbar(aes(ymin = mean_egfr - se_egfr, ymax = mean_egfr + se_egfr),
                width = 0.8, colour = "grey25") +
  geom_line(colour = "grey15", linewidth = 0.9) +
  geom_point(size = 2, colour = "grey15") +
  scale_y_continuous(limits = c(30, 115)) +
  labs(title = "Mean eGFR over time (active treatment arm)",
       x = "Week", y = "eGFR (mean ± SE)") +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[Every patient, and the summary]{.panel-lab}
```{r spaghetti, fig.width=5.2, fig.height=3.9}
egfr_medians <- summarize(egfr, egfr = median(egfr), .by = week)
steep <- filter(egfr, usubjid == "MER201-0007")
drop  <- round(100 * (steep$egfr[7] / steep$egfr[1] - 1))
ggplot(egfr, aes(week, egfr, group = usubjid)) +
  geom_line(colour = pal$g2, linewidth = 0.35) +
  geom_line(data = egfr_medians, aes(group = 1), colour = pal$ink,
            linewidth = 1.1) +
  geom_line(data = steep, colour = pal$accent, linewidth = 0.8) +
  annotate("text", x = 26.4, y = min(steep$egfr), hjust = 0, family = "Menlo",
           size = 2.9, colour = pal$accent,
           label = sprintf("0007 %d%%", drop)) +
  annotate("text", x = 26.4, y = egfr_medians$egfr[7], hjust = 0,
           family = "Charter",
           size = 3.4, colour = pal$ink, label = "median") +
  scale_x_continuous(breaks = c(0, 4, 12, 26), limits = c(0, 30.5)) +
  scale_y_continuous(limits = c(30, 115)) +
  labs(x = "week", y = "eGFR") +
  theme_tufte(15)
```
:::
::::

::: {.caption}
eGFR, active treatment arm, 30 patients over 26 weeks, same y scale in both
panels. The mean is flat, and the left panel ends the story there. The right
panel keeps the macro reading (a stable band) and adds the micro one: three
patients decline more than 30%, one past 40%, in red. Means can be stable
while patients are not.
:::

::: notes
1 min. This is the argument for keeping the spaghetti plot: a means-only
figure files the nephrotoxicity signal in an appendix listing.
:::

## Rule 9 · The smallest effective difference

::: {.eyebrow}
Rule 09 [Visual Explanations, p. 73 — distinctions as subtle as possible, still clear]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Three saturated hues, one legend]{.panel-lab .bad}
```{r loudarms, fig.width=5.2, fig.height=3.9}
ggplot(prof, aes(week, pchg, colour = dose)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  labs(title = "Tumor burden by dose (MER-102)",
       x = "Week", y = "Mean % change", colour = "Dose") +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[Gray context, one accent]{.panel-lab}
```{r quietarms, fig.width=5.2, fig.height=3.9}
ggplot(prof, aes(week, pchg, group = dose)) +
  geom_line(aes(colour = dose), linewidth = 0.7) +
  scale_colour_manual(values = c(pal$g2, pal$g3, pal$accent)) +
  geom_text(data = dose_ends,
            aes(label = dose, colour = dose), hjust = 0, nudge_x = 0.6,
            family = "Charter", size = 3.4, show.legend = FALSE) +
  scale_x_continuous(breaks = c(0, 12, 26), limits = c(0, 30.5)) +
  scale_y_continuous(breaks = c(0, -10, -20)) +
  labs(title = "Tumor burden by dose (MER-102)",
       x = "week", y = "mean % change") +
  theme_tufte(15)
```
:::
::::

::: {.caption}
The proposed phase 3 dose is 20 mg, so 20 mg carries the accent and the other
arms become gray context, labeled where their lines end. The left panel
separates three groups with three saturated hues and makes the reader look up
each one; the reading is identical, only the effort changes.
:::

::: notes
0.5 min. Contrast is a budget. Two grays and one red separate three arms;
red against green alone fails for roughly 8% of men.
:::

## Rule 10 · Integrate words and data

::: {.eyebrow}
Rule 10 [The forest plot is the most Tufte figure in the CSR: a table and a graphic, fused]{}
:::

```{r forest, fig.width=11.4, fig.height=4.4}
forest$row <- rev(seq_len(nrow(forest)))
forest$focal <- forest$subgroup == "Overall"
ggplot(forest, aes(hr, row)) +
  annotate("segment", x = 1, xend = 1, y = 0.6, yend = 9.4,
           colour = pal$g2, linewidth = 0.4) +
  geom_segment(aes(x = lo, xend = hi, yend = row,
                   colour = focal), linewidth = 0.5) +
  geom_point(aes(size = focal, colour = focal)) +
  geom_text(aes(x = 0.208, label = n), family = "Menlo", size = 3.1,
            colour = pal$g3, hjust = 1) +
  geom_text(aes(x = 2.05, label = sprintf("%.2f (%.2f–%.2f)", hr, lo, hi),
                colour = focal),
            family = "Menlo", size = 3.1, hjust = 1) +
  annotate("text", x = c(0.208, 2.05), y = 9.9, label = c("n", "HR (95% CI)"),
           family = "Menlo", size = 2.9, colour = pal$muted, hjust = 1) +
  annotate("text", x = c(0.93, 1.08), y = 0.25, hjust = c(1, 0),
           family = "Menlo", size = 2.9, colour = pal$muted,
           label = c("← favors active treatment", "favors placebo →")) +
  scale_colour_manual(values = c(`TRUE` = pal$accent, `FALSE` = pal$ink)) +
  scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1.7)) +
  scale_x_log10(breaks = c(0.5, 1), limits = c(0.14, 2.05)) +
  scale_y_continuous(breaks = forest$row, labels = forest$subgroup) +
  coord_cartesian(clip = "off") +
  labs(x = "hazard ratio, log scale", y = NULL) +
  theme_tufte(15) +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.text.y = element_text(family = "Charter", size = 11,
                                   colour = pal$ink, hjust = 0),
        plot.margin = margin(18, 8, 8, 8))
```

::: {.caption}
PFS by subgroup, MER-201. Label, n, estimate and picture share each row; the
eye never leaves the line it is reading. No legend, no second panel, no page
flip to Table 14.2.4. The point sits on a log scale where a ratio belongs,
the interval is drawn rather than footnoted, and the prior platinum CI
crossing 1 is legible at a glance.
:::

::: notes
1.5 min. No junk twin here: nobody draws an HR bar chart, and the forest plot
needs no foil. Same principle at smaller scale: sparklines beside numbers in
demographic tables; a number in a sentence can carry its shape with it.
:::

## Kill · The response pie

::: {.eyebrow}
The kill list, drawn [1 of 3]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Angles, matched to a legend]{.panel-lab .bad}
```{r pie, fig.width=5.2, fig.height=3.9}
ggplot(bor, aes(x = "", y = n, fill = response)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  labs(title = "Best overall response, active treatment arm",
       fill = NULL, x = NULL, y = NULL) +
  theme_grey(15, base_family = "Helvetica") +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(size = 9.5, hjust = 0),
        legend.text = element_text(size = 8),
        legend.key.size = unit(11, "pt"))
```
:::
::: {.column width="50%"}
[Positions, sorted and labeled]{.panel-lab}
```{r borsorted, fig.width=5.2, fig.height=3.9}
bor <- bor[order(bor$n), ]
bor$response <- factor(bor$response, levels = bor$response)
bor$focal <- bor$response %in% c("Partial response", "Complete response")
ggplot(bor, aes(n, response)) +
  geom_segment(aes(x = 0, xend = n, yend = response), colour = pal$g2,
               linewidth = 0.4) +
  geom_point(aes(colour = focal), size = 2.6) +
  geom_text(aes(label = sprintf("%d  (%.0f%%)", n, 100 * n / sum(n)),
                colour = focal),
            hjust = 0, nudge_x = 0.5, family = "Menlo", size = 3.1) +
  scale_colour_manual(values = c(`TRUE` = pal$accent, `FALSE` = pal$ink)) +
  scale_x_continuous(limits = c(0, 27), breaks = NULL) +
  labs(title = "Best overall response, active treatment arm (n = 60)",
       x = NULL, y = NULL) +
  theme_tufte(15) +
  theme(axis.line = element_blank(), axis.ticks = element_blank(),
        axis.text.y = element_text(family = "Charter", size = 10.5,
                                   colour = pal$ink))
```
:::
::::

::: {.caption}
Which is larger, stable disease or partial response? The pie asks you to
compare two angles 90 degrees apart, via a legend. The dot plot answers
before the question forms, gives the exact counts, and marks the responders,
the number the study lives on, in red.
:::

::: notes
0.75 min. People read positions on a common scale well and angles badly.
That is the whole case, and it never changes.
:::

## Kill · The dual-axis overlay

::: {.eyebrow}
The kill list, drawn [2 of 3]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Two scales, one panel]{.panel-lab .bad}
```{r dualaxis, fig.width=5.2, fig.height=3.9}
ggplot(pk, aes(week)) +
  geom_line(aes(y = conc), colour = "#2166ac", linewidth = 1) +
  geom_point(aes(y = conc), colour = "#2166ac", size = 1.8) +
  geom_line(aes(y = (pchg + 25) / 25 * 700), colour = "#b2182b",
            linewidth = 1) +
  geom_point(aes(y = (pchg + 25) / 25 * 700), colour = "#b2182b",
             size = 1.8) +
  scale_y_continuous(name = "Concentration (ng/mL)",
                     sec.axis = sec_axis(~ . / 700 * 25 - 25,
                                         name = "Tumor change (%)")) +
  labs(title = "PK and efficacy over time", x = "Week") +
  theme_junk(15) +
  theme(axis.title.y.left = element_text(colour = "#2166ac"),
        axis.title.y.right = element_text(colour = "#b2182b"),
        axis.text.y.left = element_text(colour = "#2166ac"),
        axis.text.y.right = element_text(colour = "#b2182b"))
```
:::
::: {.column width="50%"}
[Two panels, one clock]{.panel-lab}
```{r twopanels, fig.width=5.2, fig.height=3.9}
pk_long <- rbind(
  data.frame(week = pk$week, val = pk$conc,
             metric = "serum concentration, ng/mL"),
  data.frame(week = pk$week, val = pk$pchg,
             metric = "mean tumor change, %")
)
pk_long$metric <- factor(pk_long$metric,
                         levels = unique(pk_long$metric))
pk_ends <- filter(pk_long, week == 26)
ggplot(pk_long, aes(week, val)) +
  geom_line(linewidth = 0.7, colour = pal$ink) +
  geom_point(data = pk_ends, size = 1.4, colour = pal$ink) +
  geom_text(data = pk_ends, aes(label = val), hjust = 0, nudge_x = 0.7,
            family = "Menlo", size = 3, colour = pal$ink) +
  facet_wrap(~metric, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = c(0, 12, 26), limits = c(0, 30.5)) +
  labs(x = "week", y = NULL) +
  theme_tufte(15) +
  theme(panel.spacing = unit(16, "pt"))
```
:::
::::

::: {.caption}
The left panel invites a correlation the data never claimed: rescale either
axis and the crossing point moves anywhere you like. The right panel keeps
the shared clock, gives each measure its own honest scale, and lets the
reader align the timings without being steered.
:::

::: notes
0.75 min. Any two monotone series can be made to cross wherever the analyst
chooses the scales. That degree of freedom is the problem.
:::

## Kill · The rainbow heatmap

::: {.eyebrow}
The kill list, drawn [3 of 3]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Hue has no order]{.panel-lab .bad}
```{r rainbow, fig.width=5.2, fig.height=3.9}
ggplot(alt_subj, aes(factor(week), usubjid, fill = fold)) +
  geom_tile() +
  scale_fill_gradientn(colours = rev(rainbow(7)), name = "ALT\nxULN") +
  labs(title = "ALT by patient and week", x = "Week", y = NULL) +
  theme_junk(15) +
  theme(axis.text.y = element_text(size = 5),
        legend.text = element_text(size = 8))
```
:::
::: {.column width="50%"}
[One hue, light to dark, sorted]{.panel-lab}
```{r singlehue, fig.width=5.2, fig.height=3.9}
peak_fold <- summarize(alt_subj, fold = max(fold), .by = usubjid)
alt_subj$usubjid <- factor(alt_subj$usubjid,
                           levels = peak_fold$usubjid[order(peak_fold$fold)])
flagged <- filter(alt_subj, week == 26,
                  usubjid %in% peak_fold$usubjid[peak_fold$fold > 2])
ggplot(alt_subj, aes(factor(week), usubjid, fill = fold)) +
  geom_tile(colour = pal$paper, linewidth = 0.4) +
  scale_fill_gradient(low = pal$g1, high = "#7a1f18", name = "ALT xULN",
                      breaks = c(1, 2, 3)) +
  geom_text(data = flagged, aes(label = sprintf("%.1f", fold)),
            family = "Menlo", size = 2.6, colour = pal$paper) +
  labs(title = "ALT by patient and week, sorted by peak", x = "week",
       y = NULL) +
  theme_tufte(15) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line = element_blank(),
        legend.position = "right",
        legend.key.size = unit(10, "pt"),
        legend.text = element_text(family = "Menlo", size = 8,
                                   colour = pal$g3),
        legend.title = element_text(family = "Menlo", size = 8,
                                    colour = pal$muted))
```
:::
::::

::: {.caption}
ALT as multiples of ULN, 24 patients, 26 weeks. Is yellow higher than green?
The rainbow forces the reader to keep re-checking the legend, and its
brightest band lands mid-scale. One hue running light to dark reads without a
key, and sorting patients by peak puts the four who crossed 2 xULN, labeled at
week 26, in one place instead of scattered.
:::

::: notes
0.75 min. Hue is not ordinal; lightness is. Sorting is the quiet half of this
fix and does as much work as the color change.
:::

## The rest of the kill list

|   Kill | Use instead | Shown at |
|--------|-------------|----------|
| Dynamite plots | Dots and medians, all patients | Rule 1 |
| Legend far from the curves | Direct labels at line ends | Rules 3, 9 |
| Grid and frame at data weight | Faint gray, or nothing | Rules 2, 7 |
| Truncated bar axes | Zero base, or the number with its CI | Rule 5 |
| Bars for ratios | A point with an interval on a log scale | the fix, rule 10 |
| 3D anything | The 2D version | never drawn, on purpose |
| Subgroup panels on mixed scales | One shared scale | Rules 6, 8 |
| "approximately ~42%" | "42.3%" | every caption |
| p-value theater | The estimate and interval, shown | Rule 10 |

: {.kill tbl-colwidths="[34,42,24]"}

::: {.small}
If a journal or a sponsor template insists on one of these, comply, and put
the Tufte version beside it in the appendix. Reviewers notice.
:::

::: notes
0.5 min. Every row is a pointer back to a pair already on screen; nothing on
this list is hypothetical.
:::

## Before a figure ships

::: {.columns}
::: {.column width="55%"}
::: {.keep}
- [**Every patient visible**, or the summary is drawn over the points.]{}
- [**Zero-based bars**; truncated line axes labeled.]{}
- [**One accent color**, spent on the finding, not the furniture.]{}
- [**Direct labels**; a legend only past five groups.]{}
- [**Shared scales** across panels and repeat figures.]{}
- [**At-risk counts** under every KM plot.]{}
- [**Sorted** by estimate, not alphabetically, unless the SAP says otherwise.]{}
- [**Type readable from the back**: ticks 12 px, labels 14 px.]{}
:::
:::
::: {.column width="45%"}
> Above all else show the data.<br>
> Maximize the data-ink ratio.<br>
> Erase non-data-ink.<br>
> Erase redundant data-ink.<br>
> Revise and edit.
>
> [VDQI, p. 105]{.cite}

::: {.small}
The fifth line does the work. Every clean panel in this deck is the second
draft of the junk panel beside it.
:::
:::
:::

::: notes
1 min. Close on "revise and edit": the before panels in this deck are real
first drafts, not strawmen; ggplot2 and SAS defaults produce them.
:::

## Sources and reuse

::: {.lede}
The three books: *The Visual Display of Quantitative Information* (1983),
*Envisioning Information* (1990), *Visual Explanations* (1997). Quotes cited
by page.
:::

<br>

::: {.small}
MER-201 and MER-102 are simulated: `R/sim_data.R` generates every dataset,
`R/theme_tufte.R` holds the palette and themes, and every figure is plain
ggplot2, dplyr, and survival. Render with `quarto render tufte-pharma.qmd`. Reuse
freely.
:::

::: notes
0.25 min. Offer the repo; the theme drops into any TFL program.
:::
