---
title: "Show every patient"
subtitle: "Tufte's rules for the figures in a clinical study report"
author: "15 minutes, for clinical statisticians"
format:
  revealjs:
    theme: [default, theme.scss]
    width: 1280
    height: 720
    margin: 0.07
    center: false
    transition: none
    slide-number: c/t
    hash: true
    embed-resources: true
    title-slide-attributes:
      data-background-color: "#faf8f2"
execute:
  echo: false
  warning: false
  message: false
fig-dpi: 192
---

```{r setup}
source("R/theme_tufte.R")
library(survival)
adtte  <- read.csv("data/adtte.csv")
tumor  <- read.csv("data/tumor_wk12.csv")
orr    <- read.csv("data/orr.csv")
labs   <- read.csv("data/lab_means.csv")
egfr   <- read.csv("data/egfr.csv")
prof   <- read.csv("data/dose_profile.csv")
forest <- read.csv("data/forest.csv")

# KM pieces used on several slides
sf <- survfit(Surv(aval, 1 - cnsr) ~ trt, data = adtte)
med <- summary(sf)$table[, "median"]
km <- data.frame(
  time  = sf$time, surv = sf$surv, ncens = sf$n.censor,
  trt   = rep(sub("trt=", "", names(sf$strata)), sf$strata)
)
km0 <- rbind(data.frame(time = 0, surv = 1, ncens = 0,
                        trt = unique(km$trt)), km)
atrisk <- do.call(rbind, lapply(c(0, 6, 12, 18, 24), function(m)
  aggregate(aval ~ trt, adtte, function(x) sum(x >= m)) |>
    transform(month = m)))
knitr::opts_chunk$set(dev = "ragg_png")
```

## Above all else show the data

::: {.eyebrow}
A working deck [MER-201 and MER-102 are invented studies. The rules are not.]{}
:::

::: {.lede}
Edward Tufte's rules for data graphics, applied to the figures a clinical
statistician actually makes: Kaplan-Meier plots, change-from-baseline plots,
lab profiles, forest plots. Six rules, a kill list, and the checks to run
before a figure goes in the CSR.
:::

```{r cover, fig.width=11.4, fig.height=3.4}
ggplot(km0, aes(time, surv, group = trt)) +
  geom_step(aes(colour = trt), linewidth = 0.7) +
  scale_colour_manual(values = c(pal$ink, pal$g2)) +
  annotate("text", x = 24.3, y = km0$surv[km0$trt == "meriplatinib 20 mg" &
             km0$time == max(km0$time[km0$trt == "meriplatinib 20 mg"])][1],
           label = "meriplatinib", hjust = 0, family = "Charter",
           size = 4, colour = pal$ink) +
  annotate("text", x = 24.3, y = km0$surv[km0$trt == "placebo" &
             km0$time == max(km0$time[km0$trt == "placebo"])][1],
           label = "placebo", hjust = 0, family = "Charter",
           size = 4, colour = pal$g3) +
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24), limits = c(0, 28.5)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) +
  labs(x = "months", y = "PFS") +
  theme_tufte(15)
```

::: {.caption}
MER-201 progression-free survival, N = 240. Two curves, two labels, five ticks.
Everything else on this slide waits for slide 4.
:::

::: notes
1 min. Frame the talk: these rules come from three books (VDQI 1983,
Envisioning Information 1990, Visual Explanations 1997). Every figure today is
from one invented phase 2 study so the numbers reconcile across slides.
:::

## The reader is a reviewer

::: {.lede}
Your figures face the busiest readers in the industry. A DSMB member with six
studies on the table. A regulator with a 400-page CSR and a PDUFA date.
Every mark that is not data costs them attention they were saving for the data.
:::

<br>

|   | Rule | Figure it fixes |
|---|------|-----------------|
| 1 | Show the data | The dynamite plot |
| 2 | Erase non-data-ink | The default Kaplan-Meier |
| 3 | Never let the picture out-run the number | The truncated ORR axis |
| 4 | Small multiples | The six-lab safety story |
| 5 | Show every patient | The spaghetti plot, kept |
| 6 | Integrate words and data | The forest plot, done right |

: {tbl-colwidths="[6,47,47]"}

::: notes
1.5 min. Tufte's frame: data-ink over total ink. In pharma the stakes are
higher than aesthetics; a figure that hides patients hides risk.
:::

## Rule 1 · Show the data

::: {.eyebrow}
Rule 01 [VDQI, p. 92]{}
:::

:::: {.columns}
::: {.column width="50%"}
[As most tools draw it]{.panel-lab .bad}
```{r dynamite, fig.width=5.2, fig.height=3.9}
dyn <- aggregate(pchg ~ trt, tumor, function(x)
  c(m = mean(x), se = sd(x) / sqrt(length(x))))
dyn <- do.call(data.frame, dyn)
ggplot(dyn, aes(trt, pchg.m, fill = trt)) +
  geom_col(width = 0.55, colour = "grey25") +
  geom_errorbar(aes(ymin = pchg.m - pchg.se, ymax = pchg.m + pchg.se),
                width = 0.18, linewidth = 0.7, colour = "grey25") +
  scale_x_discrete(labels = c("meriplatinib", "placebo")) +
  labs(title = "Mean change in tumor size, week 12",
       x = NULL, y = "% change (mean ± SE)") +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[The same 120 patients]{.panel-lab}
```{r dots, fig.width=5.2, fig.height=3.9}
set.seed(4)
med_t <- aggregate(pchg ~ trt, tumor, median)
ggplot(tumor, aes(trt, pchg)) +
  geom_hline(yintercept = 0, colour = pal$g2, linewidth = 0.3) +
  geom_jitter(width = 0.16, size = 1.1, alpha = 0.55, colour = pal$ink) +
  geom_errorbar(data = med_t, aes(ymin = pchg, ymax = pchg),
                width = 0.4, linewidth = 0.8, colour = pal$accent) +
  geom_text(data = med_t, aes(label = sprintf("%.0f%%", pchg)),
            hjust = 0, nudge_x = 0.26, family = "Menlo",
            size = 3.4, colour = pal$accent, fontface = "bold") +
  scale_x_discrete(labels = c("meriplatinib", "placebo")) +
  labs(title = "Change in tumor size, week 12, every patient",
       x = NULL, y = "% change from baseline") +
  theme_tufte(15)
```
:::
::::

::: {.caption}
The bar shows 2 numbers from 120 patients and hides the rest: the spread, the
placebo responders, the meriplatinib progressors. The dot plot shows all 120
and the medians in red. Same data, opposite amount of information.
:::

::: notes
2.5 min. The dynamite plot is the most common figure in the industry and the
least informative. A mean and an SE from 60 patients per arm. Point at the
placebo patients below -30%: invisible on the left.
:::

## Rule 2 · Erase non-data-ink

::: {.eyebrow}
Rule 02 [VDQI, p. 96 — data-ink is ink that would change if the data changed]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Default output]{.panel-lab .bad}
```{r kmbad, fig.width=5.2, fig.height=3.9}
ggplot(km0, aes(time, surv, colour = trt)) +
  geom_step(linewidth = 1.1) +
  geom_point(data = subset(km0, ncens > 0), shape = 3, size = 1.6) +
  scale_colour_manual(NULL, values = c("#00BFC4", "#F8766D")) +
  scale_x_continuous(breaks = 0:8 * 3) +
  labs(title = "Kaplan-Meier plot of PFS (FAS)",
       x = "Time (months)", y = "Survival probability") +
  theme_junk(15) +
  theme(legend.position = c(0.78, 0.85))
```
:::
::: {.column width="50%"}
[After erasure]{.panel-lab}
```{r kmgood, fig.width=5.2, fig.height=3.9}
ends <- aggregate(cbind(time, surv) ~ trt, km0, function(x) x[length(x)])
ggplot(km0, aes(time, surv, group = trt)) +
  geom_step(aes(colour = trt), linewidth = 0.7) +
  geom_point(data = subset(km0, ncens > 0), aes(colour = trt),
             shape = "|", size = 1.8) +
  scale_colour_manual(values = c(pal$ink, pal$g2)) +
  annotate("segment", x = med[1], xend = med[1], y = 0, yend = 0.5,
           colour = pal$accent, linewidth = 0.35, linetype = "22") +
  annotate("text", x = med[1] + 0.4, y = 0.06, family = "Menlo", size = 3.2,
           colour = pal$accent, hjust = 0,
           label = sprintf("median %.1f mo", med[1])) +
  annotate("text", x = ends$time[1] + 0.5, y = ends$surv[1] + 0.03,
           label = "meriplatinib", hjust = 0, family = "Charter",
           size = 3.6, colour = pal$ink) +
  annotate("text", x = ends$time[2] + 0.5, y = ends$surv[2] - 0.01,
           label = sprintf("placebo\nmedian %.1f mo", med[2]), hjust = 0,
           family = "Charter", size = 3.6, colour = pal$g3, lineheight = 0.95) +
  geom_text(data = subset(atrisk, trt == "meriplatinib 20 mg"),
            aes(month, -0.24, label = aval), family = "Menlo",
            size = 2.9, colour = pal$g3) +
  geom_text(data = subset(atrisk, trt == "placebo"),
            aes(month, -0.34, label = aval), family = "Menlo",
            size = 2.9, colour = pal$g2) +
  annotate("text", x = -1.6, y = c(-0.24, -0.34), hjust = 1, family = "Menlo",
           size = 2.9, colour = c(pal$g3, pal$g2),
           label = c("at risk  mer", "plc")) +
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24), limits = c(-3.5, 30.5)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(x = NULL, y = "PFS") +
  theme_tufte(15) +
  theme(plot.margin = margin(8, 8, 44, 8))
```
:::
::::

::: {.caption}
Erasure order: frame box, grid, legend, minor ticks. What replaced them is
data: direct labels, censor marks, medians where they occur, and the at-risk
counts, which are the honest width of the right-hand tail.
:::

::: notes
2.5 min. The legend is the tell: it asks the reader to hold color pairs in
memory. The at-risk table is not decoration; past month 18 both curves stand
on fewer than 35 patients.
:::

## Rule 3 · Never let the picture out-run the number

::: {.eyebrow}
Rule 03 [Lie factor = visual effect ÷ data effect. Keep it between 0.95 and 1.05]{}
:::

:::: {.columns}
::: {.column width="50%"}
[Axis starts at 35]{.panel-lab .bad}
```{r orrbad, fig.width=5.2, fig.height=3.6}
ggplot(orr, aes(trt, orr, fill = trt)) +
  geom_col(width = 0.5, colour = "grey25") +
  coord_cartesian(ylim = c(35, 45)) +
  scale_x_discrete(labels = c("meriplatinib", "placebo")) +
  labs(title = "Overall response rate (%)", x = NULL, y = NULL) +
  theme_junk(15)
```
:::
::: {.column width="50%"}
[Axis starts at zero]{.panel-lab}
```{r orrgood, fig.width=5.2, fig.height=3.6}
ggplot(orr, aes(trt, orr)) +
  geom_col(width = 0.5, fill = c(pal$accent, pal$g2)) +
  geom_text(aes(label = sprintf("%.1f", orr)), vjust = -0.6,
            family = "Menlo", size = 3.6, colour = pal$ink) +
  scale_x_discrete(labels = c("meriplatinib", "placebo")) +
  scale_y_continuous(limits = c(0, 48), breaks = c(0, 45)) +
  labs(title = "Overall response rate (%)", x = NULL, y = NULL) +
  theme_tufte(15)
```
:::
::::

::: {.caption}
ORR 42.3% vs 38.1%: an 11% relative difference. The truncated panel draws a
135% visual difference. Lie factor 12. Bars encode length, so their baseline
is zero; if a difference is only visible on a truncated axis, that is the
finding, and the honest home for it is a dot plot with a labeled range or the
number itself with its CI.
:::

::: notes
2 min. Also on the lie list: response "bubbles" scaled by radius, dual-axis
PK-efficacy overlays, subgroup panels on different scales.
:::

## Rule 4 · Small multiples

::: {.eyebrow}
Rule 04 [VDQI, p. 67 — "Compared to what?"]{}
:::

```{r labs, fig.width=11.4, fig.height=4.4}
labs$analyte <- factor(labs$analyte, levels = c("ALT", "AST", "ALP",
                       "Bilirubin", "Creatinine", "Hemoglobin"))
labs$key <- interaction(labs$analyte, labs$trt)
labs$col <- ifelse(labs$trt == "placebo", "plc",
                   ifelse(labs$analyte == "ALT", "focal", "act"))
lab_end <- subset(labs, week == 26 & trt != "placebo")
ggplot(labs, aes(week, pchg, group = key, colour = col)) +
  geom_hline(yintercept = 0, colour = pal$g1, linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(data = lab_end, size = 1.4) +
  scale_colour_manual(values = c(plc = pal$g2, act = pal$ink,
                                 focal = pal$accent)) +
  facet_wrap(~analyte, nrow = 2) +
  scale_x_continuous(breaks = c(0, 12, 26)) +
  scale_y_continuous(breaks = c(0, 25)) +
  labs(x = "week", y = "mean % change") +
  theme_tufte(15)
```

::: {.caption}
Mean % change from baseline, meriplatinib in ink, placebo in gray. Six panels,
one scale, same order every time this figure appears in the CSR. The ALT climb
to +24% is the safety story, so ALT carries the accent; nothing else does.
Six lines overlaid on one panel would bury it.
:::

::: notes
2 min. The eye reads parallel panels faster than it untangles overlap.
Identical scale in every panel or the comparison is a lie (see rule 3).
:::

## Rule 5 · Show every patient

::: {.eyebrow}
Rule 05 [Envisioning Information — "To clarify, add detail"]{}
:::

```{r spaghetti, fig.width=11.4, fig.height=4.2}
med_e <- aggregate(egfr ~ week, egfr, median)
steep <- subset(egfr, usubjid == "MER201-0007")
drop  <- round(100 * (steep$egfr[7] / steep$egfr[1] - 1))
ggplot(egfr, aes(week, egfr, group = usubjid)) +
  geom_line(colour = pal$g2, linewidth = 0.35) +
  geom_line(data = med_e, aes(group = 1), colour = pal$ink, linewidth = 1.1) +
  geom_line(data = steep, colour = pal$accent, linewidth = 0.8) +
  annotate("text", x = 26.4, y = min(steep$egfr), hjust = 0, family = "Menlo",
           size = 3.2, colour = pal$accent,
           label = sprintf("MER201-0007  %d%%", drop)) +
  annotate("text", x = 26.4, y = med_e$egfr[7], hjust = 0, family = "Charter",
           size = 3.6, colour = pal$ink, label = "median") +
  scale_x_continuous(breaks = c(0, 4, 12, 26), limits = c(0, 29.5)) +
  labs(x = "week", y = "eGFR") +
  theme_tufte(15)
```

::: {.caption}
eGFR, meriplatinib arm, 30 patients over 26 weeks. The median is flat. Three
patients decline more than 30%, one past 40%. A means-only figure reports the
flat line and files the nephrotoxicity signal in an appendix listing.
:::

::: notes
2 min. Micro and macro: from the back of the room, a stable band. Up close,
every patient traceable. Means can be stable while patients are not; this is
the argument for keeping the spaghetti plot.
:::

## Rule 6 · Integrate words and data

::: {.eyebrow}
Rule 06 [The forest plot is the most Tufte figure in the CSR: a table and a graphic, fused]{}
:::

```{r forest, fig.width=11.4, fig.height=4.4}
forest$row <- rev(seq_len(nrow(forest)))
forest$focal <- forest$subgroup == "Overall"
ggplot(forest, aes(hr, row)) +
  annotate("segment", x = 1, xend = 1, y = 0.6, yend = 9.4,
           colour = pal$g2, linewidth = 0.4) +
  geom_segment(aes(x = lo, xend = hi, yend = row,
                   colour = focal), linewidth = 0.5) +
  geom_point(aes(size = focal, colour = focal)) +
  geom_text(aes(x = 0.208, label = n), family = "Menlo", size = 3.1,
            colour = pal$g3, hjust = 1) +
  geom_text(aes(x = 2.05, label = sprintf("%.2f (%.2f\u2013%.2f)", hr, lo, hi),
                colour = focal),
            family = "Menlo", size = 3.1, hjust = 1) +
  annotate("text", x = c(0.208, 2.05), y = 9.9, label = c("n", "HR (95% CI)"),
           family = "Menlo", size = 2.9, colour = pal$muted, hjust = 1) +
  annotate("text", x = c(0.93, 1.08), y = 0.25, hjust = c(1, 0),
           family = "Menlo", size = 2.9, colour = pal$muted,
           label = c("\u2190 favors meriplatinib", "favors placebo \u2192")) +
  scale_colour_manual(values = c(`TRUE` = pal$accent, `FALSE` = pal$ink)) +
  scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1.7)) +
  scale_x_log10(breaks = c(0.5, 1), limits = c(0.14, 2.05)) +
  scale_y_continuous(breaks = forest$row, labels = forest$subgroup) +
  coord_cartesian(clip = "off") +
  labs(x = "hazard ratio, log scale", y = NULL) +
  theme_tufte(15) +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.text.y = element_text(family = "Charter", size = 11,
                                   colour = pal$ink, hjust = 0),
        plot.margin = margin(18, 8, 8, 8))
```

::: {.caption}
PFS by subgroup, MER-201. Label, n, estimate and picture share each row; the
eye never leaves the line it is reading. No legend, no second panel. The prior
platinum CI crossing 1 is legible without a footnote.
:::

::: notes
1.5 min. Same principle at smaller scale: sparklines beside numbers in
demographic tables; a number in a sentence can carry its shape with it.
:::

## The kill list, pharma edition

:::: {.columns}
::: {.column width="50%"}
| Kill | Use instead |
|------|-------------|
| Dynamite plots | Dots + median, all patients |
| Pie of demographics | The demographics table |
| Truncated bar axes | Zero base, or a dot plot |
| Dual-axis PK / efficacy | Two aligned panels |
| 3D anything | The 2D version |
| Rainbow lab heatmaps | Single hue, light to dark |

: {.kill}
:::
::: {.column width="50%"}
| Kill | Use instead |
|------|-------------|
| Legend far from curves | Direct labels at line ends |
| Grid at data weight | Faint gray, or none |
| Symbols at every step | Censor ticks only |
| Subgroup panels, mixed scales | One shared scale |
| "approximately ~42%" | "42.3%" |
| p-value theater | Estimate + CI, shown |

: {.kill}
:::
::::

<br>

::: {.small}
If a journal or a sponsor template insists on one of these, comply, and put
the Tufte version beside it in the appendix. Reviewers notice.
:::

::: notes
1.5 min. Quick scan, don't read every row aloud. The theme: every item hides
either patients or scale.
:::

## Before a figure ships

::: {.columns}
::: {.column width="55%"}
::: {.keep}
- [**Every patient visible**, or the summary is drawn over the points.]{}
- [**Zero-based bars**; truncated line axes labeled.]{}
- [**One accent color**, spent on the finding, not the furniture.]{}
- [**Direct labels**; a legend only past five groups.]{}
- [**Shared scales** across panels and repeat figures.]{}
- [**At-risk counts** under every KM plot.]{}
- [**Sorted** by estimate, not alphabetically, unless the SAP says otherwise.]{}
- [**Type readable from the back**: ticks 12 px, labels 14 px.]{}
:::
:::
::: {.column width="45%"}
> Above all else show the data.<br>
> Maximize the data-ink ratio.<br>
> Erase non-data-ink.<br>
> Erase redundant data-ink.<br>
> Revise and edit.
>
> [VDQI, p. 105]{.cite}

::: {.small}
The fifth line does the work. Every clean panel in this deck is the second
draft of the junk panel beside it.
:::
:::
:::

::: notes
1.5 min. Close on "revise and edit": the before panels in this deck are real
first drafts, not strawmen; ggplot2 and SAS defaults produce them.
:::

## Sources and reuse

::: {.lede}
The three books: *The Visual Display of Quantitative Information* (1983),
*Envisioning Information* (1990), *Visual Explanations* (1997). Quotes cited
by page.
:::

<br>

::: {.small}
MER-201 and MER-102 are simulated: `R/sim_data.R` generates every dataset,
`R/theme_tufte.R` holds the palette and themes, and every figure is plain
ggplot2 + survival. Render with `quarto render tufte-pharma.qmd`. Reuse
freely.
:::

::: notes
0.5 min. Offer the repo; the theme drops into any TFL program.
:::
